# Analysis of alcohol consumption data (Week 3)

In this Analysis exercise, ...

## Loading data

I start by loading the alcohol consumption data set with the `read_csv` function from the `readr` library and print the name of its variables:

```{r}
# load the readr library
library(readr)

# load the data file
alc <- read_csv("data/alc.csv")

# print the variable names
colnames(alc)

```

The data of 370 observations is combined from two data sets describing student achievement in secondary education of two Portuguese schools (https://archive.ics.uci.edu/dataset/320/student+performance). The data attributes include student grades, demographic, social and school related features and it was collected by using school reports and questionnaires. This analysis studies students' alcohol consumption, which is described by four variables in the data:

- `Dalc`: Workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
- `Walc`: Weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
- `alc_use`: Average of `Dalc` and `Walc`
- `high_use`: `TRUE` if `alc_use > 2` and `FALSE` otherwise

In this report, I will mostly focus on the last variable and its relationship with some of the other variables in the data set.

## Four variables vs. alcohol consumption

Next, I will choose four interesting variables and hypothesize about their relationship with high/low alcohol consumption. The variables and hypotheses are:

- `activities`: Extra-curricular activities (binary: yes or no). My guess is that students with extra-curricular activities have less time to consume alcohol.
- `freetime`: Free time after school (numeric: from 1 - very low to 5 - very high). My prediction is that the amount of free time a student has correlates with the alcohol consumption.
- `famrel`: Quality of family relationships (numeric: from 1 - very bad to 5 - excellent). I hypothesize that this variable has an anti-correlation with alcohol use: a high quality of family relationship promotes less drinking and vice versa.
- `studytime`: Weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours). For this one, I also assume an anti-correlation: students who study more are less likely to have time and/or motivation to go out and drink.

Let's study numerically and graphically the distributions of the chosen variables and see if the hypotheses hold. First, it is useful to check the distributions of each variable of interest:

```{r}

# access the tidyverse libraries tidyr, dplyr, ggplot2
library(tidyr); library(dplyr); library(ggplot2)

# draw a bar plot of each variable
select(alc, all_of(c("high_use","activities","freetime","famrel","studytime"))) %>% gather %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_bar()

```

It seems like almost third of the students are high users of alcohol. About half of them have extra-curricular activities, most of them have good family relations, and most study 2-5 hours a week. The `freetime` variable is not defined quantitatively, but most of the students consider to have intermediate amount of free time after school.

Next, some cross tabulations might reveal some relationships between the variables,

```{r}
# Cross tabulation of high alcohol use and activities
table(alc$high_use, alc$activities)

# Cross tabulation of high alcohol use and free time
table(alc$high_use, alc$freetime)

# Cross tabulation of high alcohol use and family relations
table(alc$high_use, alc$famrel)

# Cross tabulation of high alcohol use and study time
table(alc$high_use, alc$studytime)

```

The same information presented graphically is perhaps easier to analyze,

```{r}

# Plot high_use vs. activities
ggplot(data = alc, aes(x = high_use)) + geom_bar() + facet_wrap("activities") + ggtitle("Extra-curricular activities")

# Plot high_use vs. freetime
ggplot(data = alc, aes(x = high_use)) + geom_bar() + facet_wrap("freetime") + ggtitle("Free time after school")

# Plot high_use vs. famrel
ggplot(data = alc, aes(x = high_use)) + geom_bar() + facet_wrap("famrel") + ggtitle("Quality of family relationships")

# Plot high_use vs. studytime
ggplot(data = alc, aes(x = high_use)) + geom_bar() + facet_wrap("studytime") + ggtitle("Weekly study time")


```

In all of the four cases, the data seems to support my hypotheses at least by visually inspecting the distributions. Let's approach the analysis more systematically and try to further solidify the hypotheses by using logistic regression to find statistically significant relationships between alcohol consumption (target variable) and the four variables. 

```{r}

# find the model with glm()
m <- glm(high_use ~ activities + freetime + famrel + studytime, data = alc, family = "binomial")

# print out a summary of the model
summary(m)

```


