# Analysis of longitudinal data (Week 6)

In this Analysis exercise, I analyze two data sets containing longitudinal data.

## RATS data

RATS data set contains data from a nutrition study conducted in three groups of rats (Crowder and Hand, 1990). The three groups were put on different diets, and each animal’s body weight (grams) was recorded repeatedly (approximately weekly, except in week seven when two recordings were taken) over a 9-week period. The question of most interest is whether the growth profiles of the three groups differ.

### Loading the data

First, I load the RATS data set, convert the categorical variables to factors, and explore its structure and dimensions,

```{r}

# Load necessary libraries
library(dplyr)
library(tidyr)
library(readr)

# load the data
RATSL <- read_csv("data/RATSL.csv")

# convert the categorical variables to factors
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)

# print out structure and dimenstions
str(RATSL)

```
The relevant variables are:

- `ID`: identification number of a rat,
- `Group`: group of a rat,
- `Weight`: weight of a rat,
- `Time`: time in days.

### Plotting the data

The weights of the rats can be plotted against time,

```{r}

#Access the package ggplot2
library(ggplot2)

# Draw the plot
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")

```

This simple graph makes a number of features of the data readily apparent. First, the weight of almost all the rats is increasing over the nine weeks of the study. Second, the rats who have higher weights at the beginning tend to have higher values throughout the study. This phenomenon is generally referred to as tracking. Third, group 1 seems to have clearly lower weights throughout the study compared to groups 2 and 3.

The tracking phenomenon can be seen more clearly in a plot of the standardized values of each observation, i.e., the values obtained by subtracting the relevant occasion mean from the original observation and then dividing by the corresponding visit standard deviation. The plot with standardized data is obtained with

```{r}

# Standardise the variable Weight
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdWeight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL)

# Plot again with the standardised Weight
ggplot(RATSL, aes(x = Time, y = stdWeight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Standardized weight (grams)") +
  theme(legend.position = "top")
```

With large numbers of observations, graphical displays of individual response profiles are of little use and investigators then commonly produce graphs showing average profiles for each nutrition group along with some indication of the variation of the observations at each time point. The mean response profiles for the three groups are given by

```{r}

# Number of subjects per group:
n <- c(8,4,4)

# Summary data with mean and standard error of Weight by Group and Time 
RATSS <- RATSL %>%
  group_by(Group, Time) %>%
  summarise(mean = mean(Weight), se = sd(Weight)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSS)

# Plot the mean profiles
ggplot(RATSS, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.7) +
  theme(legend.position = c(0.8,0.45)) +
  scale_y_continuous(name = "mean(Weight) +/- se(Weight)")

```

There is considerable overlap in the mean profiles of the nutrition groups 2 and 3 is suggesting perhaps that there is little difference between the two groups with respect to the mean weights.


### Summary measure analysis

As our first example of the summary measure approach it will be applied to the values of the weights. The mean of weeks 1 to 9 will be the chosen summary measure. We first calculate this measure and then
look at boxplots of the measure for each nutrition group. The resulting plot is

```{r}

# Create a summary data by Group and ID with mean as the summary variable (ignoring baseline time 0)
RATSL8S <- RATSL %>%
  filter(Time > 1) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()

# Glimpse the data
glimpse(RATSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(RATSL8S, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time (1-8 days)")

```

The diagram indicates that the mean summary measure is most variable in the second nutrition group and its distribution in this group is somewhat skew. The boxplots of all the groups reveal an outlier. They might bias the conclusions from further comparisons of the groups, so we decide to remove those rats from the data. The new version of the boxplots is 

```{r}

# Create a new data by filtering the outliers and adjust the ggplot code the draw the plot again with the new data
RATSL8S1 <- RATSL8S %>% filter((mean > 240 & Group == 1) | (mean < 590 & Group == 2) | (mean > 500 & Group == 3))

# Draw a boxplot of the mean versus group
ggplot(RATSL8S1, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(Weight), Time (1-8 days)")

```

Without the outliers, there is clear evidence of a difference in location of the summary measure distributions in each group. The difference is mostly a result of the different baselines of each group.

Baseline measurements of the outcome variable in a longitudinal study are often correlated with the chosen summary measure and using such measures in the analysis can often lead to substantial gains in precision when used appropriately as a covariate in an analysis of covariance (see Everitt and Pickles, 2004). We can illustrate the analysis on the RATS data using the weight corresponding to time one (day) taken on the first day of the study as the baseline covariate. The results are

```{r}

# Add the baseline from the original data as a new variable to the summary data
RATSL8S2 <- RATSL8S %>%
  mutate(baseline = filter(RATSL, Time == 1)$Weight)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = RATSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

We see that the baseline weight is strongly related to the weights measured after the study has begun, but there is no evidence of a nutrition group difference even after conditioning on the baseline value.

## BPRS data

Next, I study data taken from Davis (2002). Here 40 male subjects were randomly assigned to one of two treatment groups and each subject was rated on the brief psychiatric rating scale (BPRS) measured before treatment began (week 0) and then at weekly intervals for eight weeks. The BPRS assesses the level of 18 symptom constructs such as hostility, suspiciousness, hallucinations and grandiosity; each of these is rated from one (not present) to seven (extremely severe). The scale is used to evaluate patients suspected of having schizophrenia.

### Loading the data

First, I load the BPRS data set, convert the categorical variables to factors, and explore its structure and dimensions,

```{r}

# load the data
BPRSL <- read_csv("data/BPRSL.csv")

# convert the categorical variables to factors
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)

# print out structure and dimenstions
str(BPRSL)

```

The relevant variables are:

- `treatment`: factor determining the treatment group of a subject (1 or 2),
- `subject`: idenfier for a subject,
- `bprs`: BPRS score of a subject,
- `week`: time in weeks.

### Plotting the data

To begin, we will plot the data, identifying the observations in each treatment group,

```{r}

# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))

```

The difference between the two groups does not seem to be obvious.

### Fitting a multiple linear regression model

Ignoring the repeated-measures structure of the data, we might fit a multiple linear regression model with BPRS score as response and time and treatment group as explanatory variables. Fitting the model gives the results

```{r}

# create a regression model BPRS_reg
BPRS_reg <-  lm(bprs ~ week + treatment, data = BPRSL)

# print out a summary of the model
summary(BPRS_reg)

```

As we might have anticipated from the plot, group 2 does not differ significantly from group 1 conditional on time; however, the regression on time is highly significant. We might go on to fit a model with a treatment group × time interaction, but we will not do this because we know from the structure of the data that the model considered here is wrong. The model assumes independence of the repeated measures of the BPRS score, and this assumption is highly unlikely. So, now we will move on to consider both some more appropriate graphics and appropriate models.

### Random intercept model

This model allows the linear regression fit for each subject to differ in intercept from other subjects. Fitting this model gives the results 

```{r}

# access library lme4
library(lme4)

# Create a random intercept model
BPRS_ref <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)

# Print the summary of the model
summary(BPRS_ref)

```

The estimated variance of the subject random effects is not that large, indicating the small variation in the intercepts of the regression fits of the individual subject BPRS score profiles. The estimated regression parameters for time and the one dummy variable are very similar to those from fitting the independence model shown in the previous section. In this example, the conclusions from the independence model and the random intercept model are the same, but in other examples this will not necessarily be so, as we shall see later.

Now we can move on to fit the random intercept and random slope model to the BPRS score data. The results are

```{r}

# create a random intercept and random slope model
BPRS_ref1 <- lmer(bprs ~ week + treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref1)

# perform an ANOVA test on the two models
anova(BPRS_ref1, BPRS_ref)

```

The results for the fixed effects are very similar to those in the previous analysis, but the likelihood ratio test for the random intercept model versus the random intercept and slope model gives a chi-squared statistic of 7.27 with 2 degrees of freedom (DF), and the associated p-value is somewhat small. The random intercept and slope model provides a slightly better fit for these data.

Finally, we can fit a random intercept and slope model that allows for a treatment group × time interaction. Fitting this model gives the results

```{r}

# create a random intercept and random slope model with the interaction
BPRS_ref2 <- lmer(bprs ~ week + treatment + week * treatment + (week | subject), data = BPRSL, REML = FALSE)

# print a summary of the model
summary(BPRS_ref2)

# perform an ANOVA test on the two models
anova(BPRS_ref2, BPRS_ref1)

```

The likelihood ratio test of the interaction random intercept and slope model against the corresponding model without an interaction is 3.17 with 1 DF; the associated p-value is somewhat small, and we can conclude that the interaction model provides a very slightly better fit for the BPRS score data. The estimated regression parameters for the interaction indicate that the BPRS score slopes are higher for
subjects in treatment group 2 than for subjects in group 1: on average 0.72 higher with an
approximate 95% confidence interval [CI] of [-0.08, 1.52]. Since zero is included in the confidence interval, one can not claim to have any significant difference between the two treatment groups.

We can find the fitted values from the interaction model and plot the fitted BPRS scores for each subject. The observed and fitted values are plotted as

```{r}

# draw the plot of BPRSL with the observed bprs values
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "Observed bprs")

# Create a vector of the fitted values
Fitted <- fitted(BPRS_ref2)

# Create a new column fitted to BPRSL
BPRSL <- BPRSL %>% mutate(Fitted = Fitted)

# draw the plot of BPRSL with the Fitted values of weight
ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)), name = "Fitted bprs")

```

This graphic shows that the interaction model fails to fit the observed data with high precision. Perhaps there are some nonlinear effects in the observed data not captured by the linear model.




