# Clustering and classification of Boston data set (Week 4)

In this Analysis exercise, ...

## Loading the data

First, I load the Boston data set from the MASS package and explore its structure and dimensions,

```{r}

# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)

```

There are 506 observations and 14 numeric variables that are related to housing in suburbs of Boston:

- `crim`: per capita crime rate by town,
- `zn`: proportion of residential land zoned for lots over 25,000 sq.ft.,
- `indus`: proportion of non-retail business acres per town,
- `chas`: Charles River dummy variable (= 1 if tract bounds river; 0 otherwise),
- `nox`: nitrogen oxides concentration (parts per 10 million),
- `rm`: average number of rooms per dwelling,
- `age`: proportion of owner-occupied units built prior to 1940,
- `dis`: weighted mean of distances to five Boston employment centres,
- `rad`: index of accessibility to radial highways,
- `tax`: full-value property-tax rate per $10,000,
- `ptratio`: pupil-teacher ratio by town,
- `black`: $1000(Bk−0.63)^2$ $Bk$ is the proportion of blacks by town,
- `lstat`: lower status of the population (percent),
- `medv`: median value of owner-occupied homes in $1000s.

## Graphical overview of the data

Next, I give a visual representation of the distributions of the different variables and correlations between them. Let's start by summarizing the variables,

```{r}

# print out the summary of the variables
summary(Boston)

```

A more visual way to study the distributions of the variables is to draw density plots,

```{r}

# access the tidyverse libraries tidyr, dplyr, ggplot2
library(tidyr); library(dplyr); library(ggplot2);

# plot the distributions as densities
Boston %>% gather %>% ggplot(aes(value)) + facet_wrap("key", scales = "free") + geom_density()

```

In the distribution plots, I can see that at least the distribution of `rm` is somewhat normally distributed, while the others are far from a normal distribution. Most of the distributions are positively skewed: `chas`, `crim`, `dis`, `lstat`, `medv`, `nox`, and `zn`. On the other hand, `age` and `black` are negatively skewed. Further, three distributions have double peaks: `indus`, `rad`, and `tax`.

It is not surprising that the average number of rooms per dwelling, `rm`, is quite normally distributed since 

Now that we know how the variables are distributed, let's consider the correlations between them and create a correlation matrix by using the function `cor` and rounding the resulting numbers to 2-decimal precision:

```{r}

# calculate the correlation matrix and round it
cor_matrix <- cor(Boston) %>% round(2)

# print the correlation matrix
cor_matrix

```

To illustrate the correlations between the variables in a more visual manner, I plot the above matrix using the `corrplot` function from the `corrplot` library,

```{r}

# visualize the correlation matrix
library(corrplot)
corrplot(cor_matrix, method="circle", type="upper", cl.pos="b", tl.pos = "d", tl.cex=0.6)

```

As seen in the correlation plot, 
















